Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-15 write-ups/PLDI/paper.tex
updated some of the tables
Shortened related work.

@@

There has been a lot of work on accelerating blocking irregular
algorithms using GPUs, and on the \emph{persistent threads}
programming style for long-running kernels~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/egh/CedermanT08,TPO10,BNP12,Pannotia}.  These
approaches rely on the occupancy-bound execution model, and aim to
take ownership of the GPU by scheduling as many workgroups as the occupancy bound allows, relying on fair scheduling between these
workgroups.  This means the GPU is unavailable for other tasks, and
the method is not future-proof because (as discussed throughout the
paper) the occupancy-bound execution model is not guaranteed by GPU
programming models and is unlikely to be guaranteed in practice by
future GPU platforms.

As we have demonstrated for a representative set of irregular
algorithms, our cooperative kernels model allows blocking algorithms
to be upgraded to run under a fair scheduler in a manner that, with
appropriate care, facilitates responsive multitasking.

@@

Hardware support for preemptive multitasking for \nvidia GPUs has been
proposed, enabling full preemption as well as \emph{SM-draining},
whereby workgroups occupying a symmetric multiprocessor (SM; a compute
unit using our terminology) are drained until the SM becomes free for
other tasks~\cite{DBLP:conf/isca/TanasicGCRNV14}.  A follow-up work adds the notion of SM \emph{flushing},
whereby a workgroup running on an SM can be cancelled and re-scheduled
from scratch if the workgroup has not yet committed any visible
side-effects to shared memory~\cite{DBLP:conf/asplos/ParkPM15}.  Both approaches have been evaluated
using simulators, over sets of regular GPU kernels.  State-of-the-art \nvidia GPUs targeting data centers (where multi-tasking is essential) do support
preemption, though it is not clear whether they guarantee fair scheduling~\cite{PascalWhitepaper}.

The advantages of these approaches are that they do not require manual
programmer effort, and the capability to fully preempt a workgroup
guards against a rogue long-running kernel making the GPU unavailable
for other tasks.  The drawbacks are that (a) because preemption is
enabled by default, \emph{every} kernel takes a potential performance
hit, and (b) the SM draining optimisation does not help in the case of
blocking algorithms (not considered in these works), since a blocked
workgroup cannot be drained.

Our solution does not impose a performance penalty on non-cooperative
kernels, which run in an unmodified manner, and careful placement of
our new language primitives can exploit the structure of a
cooperative kernel, enabling more efficient multitasking than
kernel-oblivious preemption.

There is scope for combining the approaches: if a
cooperative kernel is insufficiently cooperative, i.e.\ it does not
invoke $\offerkill$ enough, then hardware support for preemption could
be invoked to prevent the kernel from denying other tasks access to
the GPU indefinitely.

\paragraph{Optimizing GPU scheduling}
%
A number of works have considered how best to schedule dynamic
workloads on GPUs~\cite{DBLP:journals/tog/SteinbergerKKHKS12,DBLP:conf/ics/WuCLSV15,DBLP:conf/usenix/KatoLRI11,DBLP:journals/tog/SteinbergerKBKDS14,DBLP:journals/rts/ElliottA12,DBLP:conf/sosp/RossbachCSRW11}.  Among these, the Whippletree approach employs a persistent
megakernel to schedule multiple, dynamic tasks in a manner that aims
to best utilise GPU
resources~\cite{DBLP:journals/tog/SteinbergerKBKDS14}.  The TimeGraph
approach similarly aims to optimise scheduling of competing workloads,
in the context of OpenGL, using GPU-CPU synchronisation so that the
CPU can monitor and control the work that is scheduled on the
GPU~\cite{DBLP:conf/usenix/KatoLRI11}.  To best utilise the compute
units of a GPU for a particular kernel, SM-centric program
transformations have been proposed, which aim to direct work to
specific compute units, rather than to specific workgroups, based on
the observation that in practice multiple workgroups may share the
same compute unit, so that said workgrops might be in a good position
to exploit memory caches associated with that compute
unit~\cite{DBLP:conf/ics/WuCLSV15}.  None of these approaches tackle
the problem of \emph{fair} scheduling on GPUs that our work addresses, thus they do not aid in safe deployment of blocking irregular algorithms.

\paragraph{Dynamic parallelism} Recent versions of CUDA and OpenCL allow a kernel to spawn further kernels~\cite{cuda-75,opencl2Spec}.
This \emph{dynamic parallelism} can be exploited to implement a
GPU-based scheduler, by having an initial scheduler kernel repeatedly
spawn further kernels as required, according to some scheduling
policy~\cite{DBLP:conf/ppopp/Muyan-OzcelikO16}.  However, a kernel
that uses dynamic parallelism is still subject to the scheduling
limitations of current GPU programming models, and in particular the
problem that workgroups are not fairly scheduled.  Hence this feature
does not offer a solution to the problem of deploying blocking
algorithms on GPUs in a safe manner that we tackle with this work.

@@

Cooperative multitasking was offered in older operating systems
(e.g. Windows before the 95 edition) and is still used by some
operating systems, such as RISC
OS.\footnote{\url{http://www.riscos.info/index.php/Preemptive_multitasking}}
Many programming languages offer cooperative multitasking via
variations of coroutines, e.g. coroutine objects in
Python\footnote{\url{https://docs.python.org/3/reference/datamodel.html\#coroutine-objects}}
or Lua,\footnote{\url{https://www.lua.org/pil/9.1.html}} or Ruby's
\emph{fiber}.\footnote{\url{http://ruby-doc.org/core-2.1.1/Fiber.html}}

In the context of GPUs, cooperative multitasking has been used to
refer to the standard setup today, whereby the onus is on a GPU kernel
to complete execution within a reasonable time budget and return
control to the CPU application~\cite{adriaens2012case,CPE:CPE1722}.  In this setting, the programmer is responsible for ensuring cooperation between the host application and GPU kernel so that a kernel does not run for too long.
%
This is in contrast to our notion of cooperative kernels, which is designed specifically to support the needs of long-running kernels that one might wish to multitask with other workloads.

>>>>

\ADComment{We have to explain in this seciton how we get fairness.}

@@

There has been a lot of work on accelerating blocking irregular
algorithms using GPUs, and on the \emph{persistent threads}
programming style for long-running
kernels~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/egh/CedermanT08,TPO10,BNP12,Pannotia}.
These approaches rely on the occupancy-bound execution model, flooding
available compute units with work, so that the GPU is unavailable for
other tasks, and assuming fair scheduling between occupant workgroups,
which is unlikely to be guaranteed on future GPU platforms.
%
As our experiments demonstrate, our cooperative kernels model allows blocking algorithms
to be upgraded to run under a fair scheduler in a manner that, with
appropriate care, facilitates responsive multitasking.

@@

Hardware support for preemption has been proposed for \nvidia GPUs,
as well as \emph{SM-draining}
whereby workgroups occupying a symmetric multiprocessor (SM; a compute
unit using our terminology) are allowed to complete until the SM becomes free for
other tasks~\cite{DBLP:conf/isca/TanasicGCRNV14}.  SM draining is limited the presence of blocking constructs, since it may not be possible to drain a blocked workgroup.
%
A follow-up work adds the notion of SM \emph{flushing},
where a workgroup can be cancelled and re-scheduled
from scratch if it has not yet committed side-effects~\cite{DBLP:conf/asplos/ParkPM15}.  Both approaches have been evaluated
using simulators, over sets of regular GPU kernels.  Very recent \nvidia GPUs targeting data centers now support
preemption, though it is not clear whether they guarantee fair scheduling~\cite{PascalWhitepaper}.

The advantages of preemption are that it does not require programmer
effort, and guards against a rogue long-running kernel making the GPU
unavailable for other tasks.  The main drawback is that \emph{every}
kernel takes a potential performance hit.
%
If implemented natively, our solution does not impact performance on regular kernels that are not scheduled with cooperative kernels.
Furthermore, careful placement of
our new primitives can exploit problem structure, enabling more efficient multitasking than
application-oblivious preemption.  There is scope for combining the approaches: if a
cooperative kernel is insufficiently cooperative, i.e.\ it does not
invoke $\offerkill$ enough, then hardware support for preemption could
be invoked to prevent the kernel from denying other tasks access to
the GPU indefinitely.

\paragraph{Optimizing GPU scheduling}
%
A number of works have considered how best to schedule dynamic
workloads on GPUs~\cite{DBLP:journals/tog/SteinbergerKKHKS12,DBLP:conf/ics/WuCLSV15,DBLP:conf/usenix/KatoLRI11,DBLP:journals/tog/SteinbergerKBKDS14,DBLP:journals/rts/ElliottA12,DBLP:conf/sosp/RossbachCSRW11}.  Among these, the Whippletree approach employs a persistent
megakernel to schedule multiple, dynamic tasks in a manner that aims
to best utilise GPU
resources~\cite{DBLP:journals/tog/SteinbergerKBKDS14}, and the TimeGraph
approach similarly aims to optimise scheduling of competing workloads,
in the context of OpenGL~\cite{DBLP:conf/usenix/KatoLRI11}.
None of these approaches tackle
the problem of \emph{fair} scheduling on GPUs, thus they do not aid in safe deployment of blocking irregular algorithms.

CUDA and OpenCL provide the facility for a kernel to spawn further kernels~\cite{cuda-75,opencl2Spec}.
This \emph{dynamic parallelism} can be used to implement a
GPU-based scheduler, by having an initial scheduler kernel repeatedly
spawn further kernels as required, according to some scheduling
policy~\cite{DBLP:conf/ppopp/Muyan-OzcelikO16}.  However, kernels
that uses dynamic parallelism are still prone to unfair scheduling of workgroups, so this feature
does not help solve the problem of safely deploying blocking
algorithms on GPUs.

@@

Cooperative multitasking was offered in older operating systems
(e.g. Windows before the 95 edition) and is still used by some
operating systems, such as RISC
OS.\footnote{\url{http://www.riscos.info/index.php/Preemptive_multitasking}}
Many programming languages offer cooperative multitasking via
variations of coroutines, e.g. coroutine objects in
Python
%\footnote{\url{https://docs.python.org/3/reference/datamodel.html\#coroutine-objects}}
and Lua,
%\footnote{\url{https://www.lua.org/pil/9.1.html}}
and Ruby's
\emph{fiber}.
%\footnote{\url{http://ruby-doc.org/core-2.1.1/Fiber.html}}

In the context of GPUs, cooperative multitasking has been used to
refer to the standard programming model, where the onus is on the GPU kernel
to complete execution within a reasonable time budget~\cite{adriaens2012case,CPE:CPE1722}.
%
This is in contrast to our cooperative kernels, which specifically aim to support the needs of long-running GPU tasks.
